%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter4.tex
%% NOVA thesis document file
%%
%% Chapter with lots of dummy text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter4.tex}%

\chapter{Work Plan}
\label{cha:work_plan}

%Grant started on 1st december 2019, 42 months.
%So originally ends at 1st june 2023. Add another 6 months. 1st december 2023
%So another 2-3 months of covid... 1st march 2024?

In this chapter we describe how we plan to use the remaining time until the end of the PhD.
The challenges and motivation of each remaining task have been covered in Chapter~\ref{cha:research_statement}; here we focus on their scheduling, dependencies, and expected dissemination.
Figure~\ref{fig:work_plan} shows the planned distribution of time across the remaining tasks.

%\begin{center}
%	\begin{tabular}{c | c}
%		\hline
%		Time frame & Activity \\
%		\hline
%		June 2022 to September 2022 & garbage collection for version management \\
%		September 2022 to November 2022 & 
%		Implementation of alternative version management solution \\
%		September 2022 to January 2023 & Evaluation of garbage collection and version management solutions \\
%		November 2022 to May 2023 & Dynamic partitioning \\
%		March 2023 to June 2023 & Evaluation of PotionDB with dynamic partitioning \\
%		February 2023 to November 2023 & Consistency levels \\
%		July 2023 to December 2023 & Evaluation of consistency levels + final PotionDB evaluation \\
%		September 2023 to March 2024 & Writing of PhD thesis and final publications \\
%		\hline
%	\end{tabular}
%	\label{table:work_plan}
%\end{center}

\begin{figure}[h!]
	\centering
	\resizebox{\textwidth}{!}{%
	\begin{ganttchart}[
		hgrid,
		vgrid,
		x unit=0.55cm,
		y unit chart=0.55cm,
		y unit title=0.7cm,
		title height=1,
		bar height=0.6,
		bar/.append style={fill=blue!40, draw=blue!60!black},
		bar incomplete/.append style={fill=orange!40, draw=orange!70!black},
		group/.append style={fill=gray!50, draw=black},
		title label font=\footnotesize\bfseries,
		bar label font=\footnotesize,
		group label font=\footnotesize\bfseries,
		milestone label font=\footnotesize\itshape
	]{1}{15}
		\gantttitle{2025}{6} \gantttitle{2026}{9} \\
		\gantttitlelist{7,...,12}{1} \gantttitlelist{1,...,9}{1} \\
		\ganttbar{VLDB revision}{1}{2} \\
		\ganttbar{Invariants + PotionDB SQL}{1}{3} \\
		\ganttbar{Dynamic Partitioning}{3}{7} \\
		\ganttbar{Consistency Levels}{5}{7} \\
		\ganttbar{GC + alt.\ Version Mgmt.}{8}{9} \\
		\ganttbar{Ph.D.\ Proposal preparation}{9}{10} \\
		\ganttbar{Final evaluation \& dissertation}{10}{15} \\
		\ganttmilestone{Ph.D.\ Proposal}{10} \\
		\ganttmilestone{Dissertation submission}{15}
	\end{ganttchart}}
	\caption{Tentative timeline of activities. Months are numbered from July 2025 (1) to September 2026 (15). Overlaps reflect the dependencies discussed below.}
	\label{fig:work_plan}
\end{figure}

\paragraph{Description of the planned activities.}
The remaining work is organized as follows:
\begin{description}
	\item[VLDB submission and revision.] We submitted a paper to VLDB with our core contributions to date (replication algorithm, transactional algorithm, materialized views over partial and non-uniform replication, and the version management algorithm), which is currently under review. Time is reserved to incorporate reviewer feedback in the event of a revise-and-resubmit decision.
	\item[Invariants and PotionDB SQL.] Design and prototype of (i) novel invariant-preservation mechanisms targeted at uniqueness, disjunction and view-level invariants beyond the bounded counter already supported, and (ii) a SQL-like front-end for PotionDB compatible with key--value access and view specification. The work yields preliminary results that constrain the design space of the consistency-levels task and informs the SQL/view interaction.
	\item[Dynamic partitioning.] Extension of the existing static-bucket protocol to support node leave/recovery, efficient data transfer (relevant-object-only transfers and operation-history compaction), and online repartitioning while preserving PSI+cset and view--object consistency. A first version targets the current consistency options before integrating the additional levels.
	\item[Consistency levels.] Per-transaction consistency selection in the spirit of red-blue consistency~\cite{li2012making}, ensuring that read-only transactions can choose weaker levels for performance while update transactions remain atomic so as not to break view--object consistency or invariants.
	\item[Garbage collection and alternative version management.] Replacement of the current conservative GC with a tracking-based mechanism that determines the oldest version still required by ongoing transactions, and full integration of the previously implemented Cure-style version manager~\cite{cure} for comparative evaluation.
	\item[Final evaluation and writing.] Closing comparative evaluation of PotionDB on Grid'5000~\cite{Grid5000} (TPC-H~\cite{tpch}, YCSB~\cite{ycsb} and microbenchmarks), preparation of the PhD Thesis Proposal and writing of the dissertation.
\end{description}

\paragraph{Evaluation infrastructure.}
A substantial part of the evaluation infrastructure is already in place: the full TPC-H benchmark~\cite{tpch} (dataset, updates and queries), microbenchmarks, and the YCSB benchmark~\cite{ycsb} are implemented in PotionDB, with flexibility to emulate alternative client/server configurations.
This infrastructure underpinned the VLDB submission and is reusable for subsequent tasks, which justifies the relatively short evaluation periods in Figure~\ref{fig:work_plan}.
We run our experiments on the Grid'5000 testbed~\cite{Grid5000}, which provides multiple machines for distributed experiments, and simulate geo-distribution latency with tc~\cite{tc} based on real measurements~\cite{AWSLatency}.
%A lot of work has already been done in the context of evaluating PotionDB, namely by implementing TPC-H's benchmark and other test clients.
%We believe this will make the next evaluating steps quicker and smoother, as the testing setup can be leveraged on for the next steps.
%Our evaluation will be done in the Grid'5000 testbed, which allows to make usage of multiple machines to run instances of PotionDB on.
%Latency from geo-distribution can be simulated by imposing latency on the links, which was already done in our experiments.
%
%\todo{Do I need to make any special reference to Grid 5000?}

\paragraph{Task dependencies.}
There is considerable overlap between Dynamic Partitioning and Consistency Levels.
We will first focus on Dynamic Partitioning under PotionDB's current consistency options, and then revisit it once the new consistency levels are designed: introducing stronger guarantees (in particular, strong consistency) may require additional coordination during repartitioning, so the protocol must be revisited rather than designed once and frozen.
Similarly, Invariants and PotionDB SQL precede Consistency Levels because invariant support and the SQL surface constrain which consistency levels are useful and how they should be exposed to the programmer.

%There is some considerable overlap between dynamic partitioning and consistency levels.
%This happens as supporting different consistency models and/or invariants may have impacts on how dynamic partitioning works, thus we need to make sure our partitioning algorithm can cope with different consistency guarantees.

%Results obtained during the course of this work will be published in top-venue conferences.
%At the moment, a submission to VLDB is being prepared, hence the extensive evaluation work already done.
%We plan to do other submissions to other conferences.
%Interesting publication targets besides VLDB include, for example, OSDI, SOSP, EuroSys, PODC and DISC.

\paragraph{Publication plan.}
Results obtained during the course of this work will be submitted to top-venue conferences.
Beyond the VLDB submission already in review, we plan at least one further submission centred on the dynamic-partitioning and consistency-levels contributions, and a third submission focused on invariants and view-level invariants under weak consistency.
Candidate venues include VLDB, SIGMOD, OSDI, SOSP, EuroSys, PODC and DISC.


%Self TODO: justify small testing periods with all the preparation already done with TPC-H to support testing